---
title: "SOC10_FINALPROJECT_CODE by Celine Qin"
output:
  pdf_document: default
  html_document: default
date: "2026-06-07"
---

Get coordinates for the 46 public high school addresses (32 for San Jose, 14 for Stockton). School addresses were collected on Excel sheets, manually cleaned, and available through the CA Department of Education School Directory. 
```{r coordinates, echo = TRUE, eval = FALSE, message = FALSE}
library(readxl)
library(tidygeocoder)
library(dplyr)
library(readr)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# pull in San Jose school list; keep only public high schools
sj_schools <- read_excel(
  file.path(DATA, "San Jose_CDESchoolDirectoryExport-2.xlsx"), skip = 4
) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, `Street Address`, `Street City`, `Street Zip`) %>%
  mutate(City = "San Jose")

# pull in Stockton school list; keep only public high schools
st_schools <- read_excel(
  file.path(DATA, "Stockton_CDESchoolDirectoryExport-3.xlsx"), skip = 4
) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, `Street Address`, `Street City`, `Street Zip`) %>%
  mutate(City = "Stockton")

# stack both cities into one table; build full address string for each school
all_schools <- bind_rows(sj_schools, st_schools) %>%
  mutate(Full_Address = paste(`Street Address`, `Street City`, "CA", `Street Zip`))

# send all addresses to Census batch geocoder 
geocoded <- all_schools %>%
  geocode(Full_Address, method = "census", lat = lat, long = lng)

# OpenStreetMap  when needed
failed <- which(is.na(geocoded$lat))
if (length(failed) > 0) {
  retry <- geocoded[failed, ] %>%
    geocode(Full_Address, method = "osm", lat = lat2, long = lng2)
  geocoded$lat[failed] <- retry$lat2
  geocoded$lng[failed] <- retry$lng2
}

# check
cat("Geocoded:", sum(!is.na(geocoded$lat)), "/", nrow(geocoded), "schools\n")

# save to csv
school_coords <- geocoded %>%
  select(City, School, Full_Address, lat, lng)

write_csv(school_coords, file.path(DATA, "school_coords.csv"))
cat("Saved school_coords.csv\n")
```

Create 1 mile buffer zones around each school address, 46 total across Stockton and San Jose. This buffer zone, also known as the school exposure zone, is the standard Unit of Analysis for this project.
```{r buffers, echo = TRUE, eval = FALSE, message = FALSE}
library(readr)
library(sf)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# load school coordinates 
school_coords <- read_csv(file.path(DATA, "school_coords.csv"))

# get actual map points
schools_sf <- school_coords %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326)

# CA coordinate system 
schools_sf <- st_transform(schools_sf, crs = 3310)

# draw a 1-mile buffer zone around each school
schools_buffer <- st_buffer(schools_sf, dist = 1609.34)

cat("Buffers created:", nrow(schools_buffer), "\n")

```

Get the estimated population within each 1 mile buffer zone using ACS. For use later in variable calculations.
```{r population, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(openxlsx)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# plug in Census API key
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# grab 2019-2023 ACS total population data for every census tract in Santa Clara (San Jose) and San Joaquin (Stockton) counties
pop_data <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "CA",
  county = c("Santa Clara", "San Joaquin"),
  geometry = TRUE,
  year = 2023
) %>%
  st_transform(crs = 3310)

# rebuild the school buffers so this chunk can run on its own
school_coords <- read_csv(file.path(DATA, "school_coords.csv"))
schools_buffer <- school_coords %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>%
  st_buffer(dist = 1609.34)

# for each school buffer, figure out how much of each census tract falls inside it,
# then use that fraction to estimate the population living in the buffer
pop_in_buffer <- function(buf) {
  hits <- st_intersects(pop_data, buf, sparse = FALSE)[, 1]
  tracts <- pop_data[hits, ]
  if (nrow(tracts) == 0) return(0)
  intersection <- st_intersection(tracts, buf)
  if (nrow(intersection) == 0) return(0)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  round(sum(intersection$estimate * proportions, na.rm = TRUE))
}

# run for all 46 schools
schools_buffer$population_1mi <- vapply(
  seq_len(nrow(schools_buffer)),
  function(i) pop_in_buffer(schools_buffer[i, ]),
  numeric(1)
)

# pull out needed data and save
results <- schools_buffer %>%
  st_drop_geometry() %>%
  select(City, School, population_1mi)

write.xlsx(results, file.path(DATA, "Population_Buffer_Results.xlsx"))
cat("Saved Population_Buffer_Results.xlsx\n")
```

For each of the high schools, get the % of students who did not meet the ELA and Math Standard for the 2024-2025 CAASPP Exam. Pull from the Santa Clara and San Joaquin County data sets, available through the CAASPP website.
```{r CAASPP, echo = TRUE, eval = FALSE, message = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(readxl)
library(openxlsx)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# process a CAASPP txt file and get % not meeting standard per school
get_caaspp_scores <- function(filepath) {
  read_delim(filepath, delim = "^", show_col_types = FALSE) %>%
    # keep only grade 11 (high school), all students combined, school-level rows
    filter(
      Grade == 11,
      `Student Group ID` == 1,
      !is.na(`School Name`)
    ) %>%
    # convert the % met and above column to numeric (suppressed values become NA)
    mutate(
      pct_met_above   = suppressWarnings(as.numeric(`Percentage Standard Met and Above`)),
      # % not meeting standard = everyone who didn't meet OR exceed the standard
      pct_not_meeting = 100 - pct_met_above
    ) %>%
    # average ELA (Test ID 1) and Math (Test ID 2) together per school
    group_by(`School Name`) %>%
    summarise(pct_unsatisfactory = round(mean(pct_not_meeting, na.rm = TRUE), 2),
              .groups = "drop")
}

# process both county files
sc_scores <- get_caaspp_scores(file.path(DATA, "SantaClaraCounty_CAASPPScores.txt"))
sj_scores <- get_caaspp_scores(file.path(DATA, "SanJoaquinCounty_CAASPPScores.txt"))

# load the school lists to match against
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
sj_schools <- school_coords %>% filter(City == "San Jose")
st_schools  <- school_coords %>% filter(City == "Stockton")

# match scores to 32 San Jose high schools
sj_matched <- sj_schools %>%
  left_join(sc_scores, by = c("School" = "School Name")) %>%
  select(School, pct_unsatisfactory)

# match scores to 14 Stockton high schools
st_matched <- st_schools %>%
  left_join(sj_scores, by = c("School" = "School Name")) %>%
  select(School, pct_unsatisfactory)

cat("San Jose matched:", sum(!is.na(sj_matched$pct_unsatisfactory)), "/ 32\n")
cat("Stockton matched:", sum(!is.na(st_matched$pct_unsatisfactory)), "/ 14\n")

# insert into CAASPP spreadsheet
write_caaspp <- function(xlsx_path, matched_df) {
  wb <- loadWorkbook(xlsx_path)
  df <- read_excel(xlsx_path)
  for (i in seq_len(nrow(df))) {
    school <- df$School[i]
    if (is.na(school)) next
    score <- matched_df$pct_unsatisfactory[matched_df$School == school]
    if (length(score) == 1) {
      writeData(wb, sheet = 1, x = score,
                startCol = 3, startRow = i + 1, colNames = FALSE)
    }
  }
  saveWorkbook(wb, xlsx_path, overwrite = TRUE)
  cat("Saved:", basename(xlsx_path), "\n")
}

write_caaspp(file.path(DATA, "San Jose_CAASPP.xlsx"),  sj_matched)
write_caaspp(file.path(DATA, "Stockton_CAASPP.xlsx"), st_matched)
```

Join data sets together to get two comprehensive data sets, one each for Stockton and San Jose, displaying data for every 1 mile School Exposure Buffer Zone (Unit of Analysis) in these each of these cities. For each data set, each row looks at 1 high school (and within their 1 mile buffer). Information includes relevant school information (name, district, address, school type (magnet, charter, public Y/N), grade levels), school-based transportation rating, transit distance score, vehicle access score, Mobility Access Index (average of school-based transportation rating, transit distance score, and vehicle access score), crime safety score, safety from crash score, Transportation Safety Index (average of crime safety score and safety from crash score), Risk Score (1.0 minus the average of Mobility Access Index and Transportation Safety Index), and the decimal percent 0.0-1.0 of Unsatisfactory CAASPP scores.
```{r all data, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get vehicle access score with ACS

#ACS household vehicle data for both counties
vehicle_raw <- get_acs(
  geography = "tract",
  variables = c(total_hh = "B08201_001", no_vehicle = "B08201_002"),
  state = "CA",
  county = c("Santa Clara", "San Joaquin"),
  geometry = TRUE,
  year = 2023,
  output = "wide"
) %>%
  st_transform(crs = 3310)

# rebuild 1-mile school buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
schools_buffer <- school_coords %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>%
  st_buffer(dist = 1609.34)

# per buffer, calculate weighted % of households with no vehicle
pct_no_veh_in_buffer <- function(buf) {
  hits <- st_intersects(vehicle_raw, buf, sparse = FALSE)[, 1]
  tracts <- vehicle_raw[hits, ]
  if (nrow(tracts) == 0) return(NA_real_)
  intersection <- st_intersection(tracts, buf)
  if (nrow(intersection) == 0) return(NA_real_)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  wt_hh <- intersection$total_hhE * proportions
  wt_nv <- intersection$no_vehicleE * proportions
  if (sum(wt_hh, na.rm = TRUE) == 0) return(NA_real_)
  round((sum(wt_nv, na.rm = TRUE) / sum(wt_hh, na.rm = TRUE)) * 100, 4)
}

schools_buffer$pct_no_vehicle <- vapply(
  seq_len(nrow(schools_buffer)),
  function(i) pct_no_veh_in_buffer(schools_buffer[i, ]),
  numeric(1)
)

# vehicle access score: 1 - % no vehicle / 100  (higher score = more households have cars)
vehicle_scores <- schools_buffer %>%
  st_drop_geometry() %>%
  mutate(vehicle_access_score = round(1 - (pct_no_vehicle / 100), 4)) %>%
  select(City, School, vehicle_access_score)

# load all spreadsheets

# school info (name, district, address, school type, grade levels)
sj_info <- read_excel(file.path(DATA, "San Jose_CDESchoolDirectoryExport-2.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, District, `Street Address`, `Public Yes/No`,
         `Magnet Yes/No`, `Charter Yes/No`, `Low Grade`, `High Grade`) %>%
  mutate(City = "San Jose")

st_info <- read_excel(file.path(DATA, "Stockton_CDESchoolDirectoryExport-3.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, District, `Street Address`, `Public Yes/No`,
         `Magnet Yes/No`, `Charter Yes/No`, `Low Grade`, `High Grade`) %>%
  mutate(City = "Stockton")

school_info <- bind_rows(sj_info, st_info)

# school-based transportation ratings
transport <- bind_rows(
  read_excel(file.path(DATA, "San Jose_School-based_Transportation.xlsx")) %>%
    select(School, transport_rating = `School Based Transport Rating`),
  read_excel(file.path(DATA, "Stockton_School-based_Transportation.xlsx")) %>%
    select(School, transport_rating = `School Based Transport Rating`)
)

# transit distance scores
transit <- bind_rows(
  read_excel(file.path(DATA, "SanJose_DistanceFromTransitStop.xlsx")) %>%
    select(School, transit_score = `Transit Distance Score (1.0 - Row D). 0.0 = farthest, 1.0 = closest`),
  read_excel(file.path(DATA, "Stockton_DistanceFromTransitStop.xlsx")) %>%
    select(School, transit_score = `Transit Distance Score (1.0 - Row D). 0.0 = farthest, 1.0 = closest`)
)

# crime safety scores
crime <- bind_rows(
  read_excel(file.path(DATA, "San Jose_#CrimeEvents.xlsx")) %>%
    select(School, crime_safety_score = `Crime Safety Score (0.0 = highest crime rate, 1.0 = lowest crime rate), min-max normalization inverted`),
  read_excel(file.path(DATA, "Stockton_#CrimeEvents.xlsx")) %>%
    select(School, crime_safety_score = `Crime Safety Score (0.0 = highest crime rate, 1.0 = lowest crime rate), min-max normalization inverted`)
)

# crash safety scores
crash <- bind_rows(
  read_excel(file.path(DATA, "San Jose_Crashes.xlsx")) %>%
    select(School, crash_safety_score = `Safety From Crash Score (1.0- Column D Value) (0.0 = unsafe, 1.0 = safe)`),
  read_excel(file.path(DATA, "Stockton_Crashes.xlsx")) %>%
    select(School, crash_safety_score = `Safety From Crash Score (1.0- Column D Value) (0.0 = unsafe, 1.0 = safe)`)
)

# CAASPP unsatisfactory scores (0-1 scale)
caaspp <- bind_rows(
  read_excel(file.path(DATA, "San Jose_CAASPP.xlsx")) %>%
    select(School, caaspp_unsatisfactory = `CAASPP Unsatisfactory (0-1)`),
  read_excel(file.path(DATA, "Stockton_CAASPP.xlsx")) %>%
    select(School, caaspp_unsatisfactory = `CAASPP Unsatisfactory (0-1)`)
)

# join everything together by school name

all_data <- school_info %>%
  left_join(transport,      by = "School") %>%
  left_join(transit,        by = "School") %>%
  left_join(vehicle_scores, by = c("City", "School")) %>%
  left_join(crime,          by = "School") %>%
  left_join(crash,          by = "School") %>%
  left_join(caaspp,         by = "School")

#calculate the three indices

all_data <- all_data %>%
  mutate(
    # average of school transport rating, transit score, vehicle access score
    mobility_access_index = round(
      rowMeans(cbind(transport_rating, transit_score, vehicle_access_score), na.rm = TRUE), 4),

    # average of crime safety score and crash safety score
    transportation_safety_index = round(
      rowMeans(cbind(crime_safety_score, crash_safety_score), na.rm = TRUE), 4),

    # 1 minus the average of both indices; higher = more at-risk
    risk_score = round(
      1 - rowMeans(cbind(mobility_access_index, transportation_safety_index), na.rm = TRUE), 4)
  )

# split into San Jose and Stockton

sj_data <- all_data %>% filter(City == "San Jose")
st_data <- all_data %>% filter(City == "Stockton")

cat("San Jose schools:", nrow(sj_data), "\n")
cat("Stockton schools:", nrow(st_data), "\n")
cat("Columns:", paste(names(all_data), collapse = " | "), "\n")

head(sj_data)
head(st_data)
```

FOR SAN JOSE: Choropleth Map of MAI vs. TAI.
```{r SJ - MAI vs. TAI choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# get zip codes for each San Jose school from the CDE directory
sj_zips <- read_excel(file.path(DATA, "San Jose_CDESchoolDirectoryExport-2.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, zip = `Street Zip`) %>%
  mutate(zip = substr(zip, 1, 5))

# join MAI and TSI values from sj_data
sj_school_values <- sj_data %>%
  select(School, mobility_access_index, transportation_safety_index) %>%
  left_join(sj_zips, by = "School")

# average MAI and TSI by zip code — some zips have more than one school
sj_by_zip <- sj_school_values %>%
  group_by(zip) %>%
  summarise(
    avg_mai = round(mean(mobility_access_index,       na.rm = TRUE), 4),
    avg_tsi = round(mean(transportation_safety_index, na.rm = TRUE), 4),
    .groups = "drop"
  )

# download zip code boundaries and keep only San Jose school zip codes
sj_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% sj_by_zip$zip) %>%
  st_transform(crs = 4326)

# join averaged scores to zip code boundaries
sj_map <- sj_zctas %>%
  left_join(sj_by_zip, by = c("ZCTA5CE20" = "zip"))

# MAI Choropleth
mai_min <- min(sj_map$avg_mai, na.rm = TRUE)
mai_max <- max(sj_map$avg_mai, na.rm = TRUE)

ggplot(sj_map) +
  geom_sf(aes(fill = avg_mai), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#4B0082", high = "#F5B8FF",
    name = "MAI Score",
    breaks = c(mai_min, mai_max),
    labels = c(paste0("Least (", round(mai_min, 2), ")"),
               paste0("Most (", round(mai_max, 2), ")"))
  ) +
  labs(
    title    = "Mobility Access Index (MAI) by Zip Code — San Jose Public High Schools",
    subtitle = "Darker = Less Mobility Access  |  Lighter = More Mobility Access",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )

# TSI Choropleth
tsi_min <- min(sj_map$avg_tsi, na.rm = TRUE)
tsi_max <- max(sj_map$avg_tsi, na.rm = TRUE)

ggplot(sj_map) +
  geom_sf(aes(fill = avg_tsi), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#4B0082", high = "#F5B8FF",
    name = "TSI Score",
    breaks = c(tsi_min, tsi_max),
    labels = c(paste0("Least (", round(tsi_min, 2), ")"),
               paste0("Most (", round(tsi_max, 2), ")"))
  ) +
  labs(
    title    = "Transportation Safety Index (TSI) by Zip Code — San Jose Public High Schools",
    subtitle = "Darker = Less Safe  |  Lighter = More Safe",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR SAN JOSE: Choropleth Map of Risk vs. CAASPP
```{r SJ - risk vs. CAASPP choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# get zip codes for each San Jose school from the CDE directory
sj_zips <- read_excel(file.path(DATA, "San Jose_CDESchoolDirectoryExport-2.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, zip = `Street Zip`) %>%
  mutate(zip = substr(zip, 1, 5))

# join risk score and CAASPP values from sj_data
sj_school_values <- sj_data %>%
  select(School, risk_score, caaspp_unsatisfactory) %>%
  left_join(sj_zips, by = "School")

# average by zip code, some zips have more than one school
sj_by_zip <- sj_school_values %>%
  group_by(zip) %>%
  summarise(
    avg_risk   = round(mean(risk_score,           na.rm = TRUE), 4),
    avg_caaspp = round(mean(caaspp_unsatisfactory, na.rm = TRUE), 4),
    .groups = "drop"
  )

# download 2020 zip code boundaries and keep only San Jose school zip codes
sj_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% sj_by_zip$zip) %>%
  st_transform(crs = 4326)

# join averaged scores to zip code boundaries
sj_map <- sj_zctas %>%
  left_join(sj_by_zip, by = c("ZCTA5CE20" = "zip"))

# Risk Score Choropleth; darker = more at risk
risk_min <- min(sj_map$avg_risk, na.rm = TRUE)
risk_max <- max(sj_map$avg_risk, na.rm = TRUE)

ggplot(sj_map) +
  geom_sf(aes(fill = avg_risk), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#F5B8FF", high = "#4B0082",
    name = "Risk Score",
    breaks = c(risk_min, risk_max),
    labels = c(paste0("Least Risk (", round(risk_min, 2), ")"),
               paste0("Most Risk (", round(risk_max, 2), ")"))
  ) +
  labs(
    title    = "Risk Score by Zip Code — San Jose Public High Schools",
    subtitle = "Darker = Higher Risk  |  Lighter = Lower Risk",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )

# CAASPP Choropleth; darker = more unsatisfactory
caaspp_min <- min(sj_map$avg_caaspp, na.rm = TRUE)
caaspp_max <- max(sj_map$avg_caaspp, na.rm = TRUE)

ggplot(sj_map) +
  geom_sf(aes(fill = avg_caaspp), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#F5B8FF", high = "#4B0082",
    name = "% Unsatisfactory",
    breaks = c(caaspp_min, caaspp_max),
    labels = c(paste0("Best (", round(caaspp_min * 100, 1), "%)"),
               paste0("Worst (", round(caaspp_max * 100, 1), "%)"))
  ) +
  labs(
    title    = "CAASPP Unsatisfactory Rate by Zip Code — San Jose Public High Schools",
    subtitle = "Darker = More Students Not Meeting Standard  |  Lighter = Fewer Students Not Meeting Standard",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR SAN JOSE: Choropleth Map Income
```{r SJ - Income Choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(tidycensus)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get zip codes for each San Jose school from the CDE directory
sj_zips <- read_excel(file.path(DATA, "San Jose_CDESchoolDirectoryExport-2.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  mutate(zip = substr(`Street Zip`, 1, 5)) %>%
  distinct(zip)

# pull median household income for each zip code (ZCTA) from the ACS
sj_income_zip <- get_acs(geography = "zcta", variables = "B19013_001", year = 2023) %>%
  filter(GEOID %in% sj_zips$zip) %>%
  select(zip = GEOID, median_income = estimate)

# download 2020 zip code boundaries and keep only San Jose school zip codes
sj_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% sj_zips$zip) %>%
  st_transform(crs = 4326)

# join income to zip code boundaries
sj_map <- sj_zctas %>%
  left_join(sj_income_zip, by = c("ZCTA5CE20" = "zip"))

# Median Household Income Choropleth; darker = lower income
inc_min <- min(sj_map$median_income, na.rm = TRUE)
inc_max <- max(sj_map$median_income, na.rm = TRUE)

ggplot(sj_map) +
  geom_sf(aes(fill = median_income), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#4B0082", high = "#F5B8FF",
    name = "Median Income",
    breaks = c(inc_min, inc_max),
    labels = c(paste0("Lowest ($", formatC(inc_min, format = "d", big.mark = ","), ")"),
               paste0("Highest ($", formatC(inc_max, format = "d", big.mark = ","), ")"))
  ) +
  labs(
    title    = "Median Household Income by Zip Code — San Jose Public High Schools",
    subtitle = "Darker = Lower Income  |  Lighter = Higher Income",
    caption  = "Source: 2019-2023 ACS (B19013_001)"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR STOCKTON: Choropleth Map of MAI vs. TAI.
```{r ST - MAI vs. TAI choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# get zip codes for each Stockton school from the CDE directory
st_zips <- read_excel(file.path(DATA, "Stockton_CDESchoolDirectoryExport-3.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, zip = `Street Zip`) %>%
  mutate(zip = substr(zip, 1, 5))

# join MAI and TSI values from st_data
st_school_values <- st_data %>%
  select(School, mobility_access_index, transportation_safety_index) %>%
  left_join(st_zips, by = "School")

# average MAI and TSI by zip code
st_by_zip <- st_school_values %>%
  group_by(zip) %>%
  summarise(
    avg_mai = round(mean(mobility_access_index,       na.rm = TRUE), 4),
    avg_tsi = round(mean(transportation_safety_index, na.rm = TRUE), 4),
    .groups = "drop"
  )

# download 2020 zip code boundaries and keep only Stockton school zip codes
st_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% st_by_zip$zip) %>%
  st_transform(crs = 4326)

# join averaged scores to zip code boundaries
st_map <- st_zctas %>%
  left_join(st_by_zip, by = c("ZCTA5CE20" = "zip"))

# MAI Choropleth; darker = less mobility access
mai_min <- min(st_map$avg_mai, na.rm = TRUE)
mai_max <- max(st_map$avg_mai, na.rm = TRUE)

ggplot(st_map) +
  geom_sf(aes(fill = avg_mai), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#7B2400", high = "#FFD5B8",
    name = "MAI Score",
    breaks = c(mai_min, mai_max),
    labels = c(paste0("Least (", round(mai_min, 2), ")"),
               paste0("Most (", round(mai_max, 2), ")"))
  ) +
  labs(
    title    = "Mobility Access Index (MAI) by Zip Code — Stockton Public High Schools",
    subtitle = "Darker = Less Mobility Access  |  Lighter = More Mobility Access",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )

# TSI Choropleth; darker = less safe
tsi_min <- min(st_map$avg_tsi, na.rm = TRUE)
tsi_max <- max(st_map$avg_tsi, na.rm = TRUE)

ggplot(st_map) +
  geom_sf(aes(fill = avg_tsi), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#7B2400", high = "#FFD5B8",
    name = "TSI Score",
    breaks = c(tsi_min, tsi_max),
    labels = c(paste0("Least (", round(tsi_min, 2), ")"),
               paste0("Most (", round(tsi_max, 2), ")"))
  ) +
  labs(
    title    = "Transportation Safety Index (TSI) by Zip Code — Stockton Public High Schools",
    subtitle = "Darker = Less Safe  |  Lighter = More Safe",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR STOCKTON: Choropleth Map of Risk vs. CAASPP
```{r ST - risk vs. CAASPP choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

# get zip codes for each Stockton school from the CDE directory
st_zips <- read_excel(file.path(DATA, "Stockton_CDESchoolDirectoryExport-3.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  select(School, zip = `Street Zip`) %>%
  mutate(zip = substr(zip, 1, 5))

# join risk score and CAASPP values from st_data
st_school_values <- st_data %>%
  select(School, risk_score, caaspp_unsatisfactory) %>%
  left_join(st_zips, by = "School")

# average by zip code
st_by_zip <- st_school_values %>%
  group_by(zip) %>%
  summarise(
    avg_risk   = round(mean(risk_score,            na.rm = TRUE), 4),
    avg_caaspp = round(mean(caaspp_unsatisfactory, na.rm = TRUE), 4),
    .groups = "drop"
  )

# download 2020 zip code boundaries and keep only Stockton school zip codes
st_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% st_by_zip$zip) %>%
  st_transform(crs = 4326)

# join averaged scores to zip code boundaries
st_map <- st_zctas %>%
  left_join(st_by_zip, by = c("ZCTA5CE20" = "zip"))

# Risk Score Choropleth; darker = more at risk
risk_min <- min(st_map$avg_risk, na.rm = TRUE)
risk_max <- max(st_map$avg_risk, na.rm = TRUE)

ggplot(st_map) +
  geom_sf(aes(fill = avg_risk), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#FFD5B8", high = "#7B2400",
    name = "Risk Score",
    breaks = c(risk_min, risk_max),
    labels = c(paste0("Least Risk (", round(risk_min, 2), ")"),
               paste0("Most Risk (", round(risk_max, 2), ")"))
  ) +
  labs(
    title    = "Risk Score by Zip Code — Stockton Public High Schools",
    subtitle = "Darker = Higher Risk  |  Lighter = Lower Risk",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )

# CAASPP Choropleth; darker = more unsatisfactory
caaspp_min <- min(st_map$avg_caaspp, na.rm = TRUE)
caaspp_max <- max(st_map$avg_caaspp, na.rm = TRUE)

ggplot(st_map) +
  geom_sf(aes(fill = avg_caaspp), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#FFD5B8", high = "#7B2400",
    name = "% Unsatisfactory",
    breaks = c(caaspp_min, caaspp_max),
    labels = c(paste0("Best (", round(caaspp_min * 100, 1), "%)"),
               paste0("Worst (", round(caaspp_max * 100, 1), "%)"))
  ) +
  labs(
    title    = "CAASPP Unsatisfactory Rate by Zip Code — Stockton Public High Schools",
    subtitle = "Darker = More Students Not Meeting Standard  |  Lighter = Fewer Students Not Meeting Standard",
    caption  = "Values averaged across schools within each zip code"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR STOCKTON: Choropleth Map Income
```{r Stockton - Income Choropleth, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(sf)
library(tigris)
library(tidycensus)
library(dplyr)
library(readr)
library(readxl)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get zip codes for each Stockton school from the CDE directory
st_zips <- read_excel(file.path(DATA, "Stockton_CDESchoolDirectoryExport-3.xlsx"), skip = 4) %>%
  filter(`Public Yes/No` == "Y", `High Grade` == "12") %>%
  mutate(zip = substr(`Street Zip`, 1, 5)) %>%
  distinct(zip)

# pull median household income for each zip code (ZCTA) from the ACS
st_income_zip <- get_acs(geography = "zcta", variables = "B19013_001", year = 2023) %>%
  filter(GEOID %in% st_zips$zip) %>%
  select(zip = GEOID, median_income = estimate)

# download 2020 zip code boundaries and keep only Stockton school zip codes
st_zctas <- zctas(cb = TRUE, year = 2020) %>%
  filter(ZCTA5CE20 %in% st_zips$zip) %>%
  st_transform(crs = 4326)

# join income to zip code boundaries
st_map <- st_zctas %>%
  left_join(st_income_zip, by = c("ZCTA5CE20" = "zip"))

# Median Household Income Choropleth; darker = lower income
inc_min <- min(st_map$median_income, na.rm = TRUE)
inc_max <- max(st_map$median_income, na.rm = TRUE)

ggplot(st_map) +
  geom_sf(aes(fill = median_income), color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low  = "#7B2400", high = "#FFD5B8",
    name = "Median Income",
    breaks = c(inc_min, inc_max),
    labels = c(paste0("Lowest ($", formatC(inc_min, format = "d", big.mark = ","), ")"),
               paste0("Highest ($", formatC(inc_max, format = "d", big.mark = ","), ")"))
  ) +
  labs(
    title    = "Median Household Income by Zip Code — Stockton Public High Schools",
    subtitle = "Darker = Lower Income  |  Lighter = Higher Income",
    caption  = "Source: 2019-2023 ACS (B19013_001)"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9,  hjust = 0.5),
    plot.caption  = element_text(size = 7,  hjust = 0.5),
    legend.position = "right"
  )
```

FOR SAN JOSE: Analyzing Risk Scores vs. how the area's income compares to the city's median household income. Top 5 most at-risk and least at-risk schools are considered. Table displays shown.
```{r SJ -risk vs. income only tables, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# San Jose city-wide median household income
sj_city_median <- get_acs(geography = "place", variables = "B19013_001",
  state = "CA", year = 2023) %>%
  filter(grepl("San Jose city", NAME)) %>% pull(estimate)
cat("San Jose city-wide median household income: $",
    formatC(sj_city_median, format = "d", big.mark = ","), "\n\n")

# get tract-level income for Santa Clara county with geometry
sj_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "Santa Clara", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild San Jose school buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
sj_buf <- school_coords %>% filter(City == "San Jose") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# calculate weighted average income within each 1-mile buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  intersection <- st_intersection(tracts, buf); if (nrow(intersection) == 0) return(NA_real_)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  round(weighted.mean(intersection$median_income, proportions, na.rm = TRUE))
}

sj_buf$median_income <- vapply(seq_len(nrow(sj_buf)),
  function(i) income_in_buffer(sj_buf[i, ], sj_income_raw), numeric(1))

# join income onto existing sj_data
sj_full <- sj_data %>%
  left_join(sj_buf %>% st_drop_geometry() %>%
    select(School, median_income),
    by = "School") %>%
  mutate(
    `Buffer Income`   = paste0("$", formatC(median_income, format = "d", big.mark = ",")),
    `vs. City Median` = ifelse(
      (median_income - sj_city_median) >= 0,
      paste0("+$", formatC(median_income - sj_city_median, format = "d", big.mark = ","), " above"),
      paste0("-$", formatC(abs(median_income - sj_city_median), format = "d", big.mark = ","), " below"))
  )

# top 5 most at-risk
cat("--- San Jose: Top 5 Most At-Risk ---\n")
print(sj_full %>% arrange(desc(risk_score)) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`))

# top 5 least at-risk
cat("\n--- San Jose: Top 5 Least At-Risk ---\n")
print(sj_full %>% arrange(risk_score) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`))
```

FOR SAN JOSE: Analyzing Risk Scores vs. how the area's income compares to the city's median household income WITH racial make-up also shown. Top 5 most at-risk and least at-risk schools are considered. Table displays shown.
```{r SJ -risk vs. income tables, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"

census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# San Jose city-wide median household income
sj_city_median <- get_acs(geography = "place", variables = "B19013_001",
  state = "CA", year = 2023) %>%
  filter(grepl("San Jose city", NAME)) %>% pull(estimate)
cat("San Jose city-wide median household income: $",
    formatC(sj_city_median, format = "d", big.mark = ","), "\n\n")

# get tract-level income for Santa Clara county with geometry
sj_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "Santa Clara", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild San Jose school buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
sj_buf <- school_coords %>% filter(City == "San Jose") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# calculate weighted average income within each 1-mile buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  intersection <- st_intersection(tracts, buf); if (nrow(intersection) == 0) return(NA_real_)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  round(weighted.mean(intersection$median_income, proportions, na.rm = TRUE))
}

sj_buf$median_income <- vapply(seq_len(nrow(sj_buf)),
  function(i) income_in_buffer(sj_buf[i, ], sj_income_raw), numeric(1))

# pull tract-level race COUNTS with geometry (B03002 = Hispanic origin by race)
sj_race_raw <- get_acs(geography = "tract",
  variables = c(total="B03002_001", white="B03002_003", black="B03002_004",
                amind="B03002_005", asian="B03002_006", pacisl="B03002_007",
                hispanic="B03002_012"),
  state = "CA", county = "Santa Clara",
  geometry = TRUE, year = 2023, output = "wide") %>%
  st_transform(crs = 3310)

# weighted racial % within each 1-mile buffer (areal interpolation)
race_in_buffer <- function(buf, race_sf) {
  hits <- st_intersects(race_sf, buf, sparse = FALSE)[, 1]
  tracts <- race_sf[hits, ]; if (nrow(tracts) == 0) return(rep(NA_real_, 7))
  inter <- st_intersection(tracts, buf); if (nrow(inter) == 0) return(rep(NA_real_, 7))
  ta <- as.numeric(st_area(tracts)); oa <- as.numeric(st_area(inter))
  prop <- pmin(oa / ta[match(inter$GEOID, tracts$GEOID)], 1)
  tot    <- sum(inter$totalE    * prop, na.rm = TRUE)
  black  <- sum(inter$blackE    * prop, na.rm = TRUE)
  hisp   <- sum(inter$hispanicE * prop, na.rm = TRUE)
  white  <- sum(inter$whiteE    * prop, na.rm = TRUE)
  asian  <- sum(inter$asianE    * prop, na.rm = TRUE)
  amind  <- sum(inter$amindE    * prop, na.rm = TRUE)
  pacisl <- sum(inter$pacislE   * prop, na.rm = TRUE)
  other  <- tot - black - hisp - white - asian - amind - pacisl
  if (tot == 0) return(rep(NA_real_, 7))
  round(c(black, hisp, white, asian, amind, pacisl, other) / tot * 100, 1)
}

race_mat <- t(vapply(seq_len(nrow(sj_buf)),
  function(i) race_in_buffer(sj_buf[i, ], sj_race_raw), numeric(7)))
sj_buf$`% Black`            <- paste0(race_mat[, 1], "%")
sj_buf$`% Hispanic`         <- paste0(race_mat[, 2], "%")
sj_buf$`% White`            <- paste0(race_mat[, 3], "%")
sj_buf$`% Asian`            <- paste0(race_mat[, 4], "%")
sj_buf$`% Native American`  <- paste0(race_mat[, 5], "%")
sj_buf$`% Pacific Islander` <- paste0(race_mat[, 6], "%")
sj_buf$`% Other`            <- paste0(race_mat[, 7], "%")

# join income + race onto existing sj_data
sj_full <- sj_data %>%
  left_join(sj_buf %>% st_drop_geometry() %>%
    select(School, median_income, `% Black`, `% Hispanic`, `% White`,
           `% Asian`, `% Native American`, `% Pacific Islander`, `% Other`),
    by = "School") %>%
  mutate(
    `Buffer Income`   = paste0("$", formatC(median_income, format = "d", big.mark = ",")),
    `vs. City Median` = ifelse(
      (median_income - sj_city_median) >= 0,
      paste0("+$", formatC(median_income - sj_city_median, format = "d", big.mark = ","), " above"),
      paste0("-$", formatC(abs(median_income - sj_city_median), format = "d", big.mark = ","), " below"))
  )

# top 5 most at-risk
cat("--- San Jose: Top 5 Most At-Risk ---\n")
print(sj_full %>% arrange(desc(risk_score)) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`,
         `% Black`, `% Hispanic`, `% White`, `% Asian`,
         `% Native American`, `% Pacific Islander`, `% Other`))

# top 5 least at-risk
cat("\n--- San Jose: Top 5 Least At-Risk ---\n")
print(sj_full %>% arrange(risk_score) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`,
         `% Black`, `% Hispanic`, `% White`, `% Asian`,
         `% Native American`, `% Pacific Islander`, `% Other`))
```

FOR STOCKTON: Analyzing Risk Scores vs. how the area's income compares to the city's median household income. Top 5 most at-risk and least at-risk schools are considered. Table displays shown.
```{r STOCKTON - risk vs. income only tables, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus); library(sf); library(dplyr); library(readr)
DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get Stockton city-wide median household income
st_city_median <- get_acs(geography = "place", variables = "B19013_001",
  state = "CA", year = 2023) %>%
  filter(grepl("Stockton city", NAME)) %>% pull(estimate)
cat("Stockton city-wide median household income: $",
    formatC(st_city_median, format = "d", big.mark = ","), "\n\n")

# get tract-level income for San Joaquin county with geometry
st_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "San Joaquin", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild Stockton school buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
st_buf <- school_coords %>% filter(City == "Stockton") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# calculate weighted average income within each 1-mile buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  intersection <- st_intersection(tracts, buf); if (nrow(intersection) == 0) return(NA_real_)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  round(weighted.mean(intersection$median_income, proportions, na.rm = TRUE))
}

st_buf$median_income <- vapply(seq_len(nrow(st_buf)),
  function(i) income_in_buffer(st_buf[i, ], st_income_raw), numeric(1))

# join income onto existing st_data
st_full <- st_data %>%
  left_join(st_buf %>% st_drop_geometry() %>%
    select(School, median_income),
    by = "School") %>%
  mutate(
    `Buffer Income`   = paste0("$", formatC(median_income, format = "d", big.mark = ",")),
    `vs. City Median` = ifelse(
      (median_income - st_city_median) >= 0,
      paste0("+$", formatC(median_income - st_city_median, format = "d", big.mark = ","), " above"),
      paste0("-$", formatC(abs(median_income - st_city_median), format = "d", big.mark = ","), " below"))
  )

# top 5 most at-risk
cat("--- Stockton: Top 5 Most At-Risk ---\n")
print(st_full %>% arrange(desc(risk_score)) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`))

# top 5 least at-risk
cat("\n--- Stockton: Top 5 Least At-Risk ---\n")
print(st_full %>% arrange(risk_score) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`))
```

FOR STOCKTON: Analyzing Risk Scores vs. how the area's income compares to the city's median household income WITH racial make-up also shown. Top 5 most at-risk and least at-risk schools are considered. Table displays shown.
```{r ST -risk vs. income tables, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus); library(sf); library(dplyr); library(readr)
DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get Stockton city-wide median household income
st_city_median <- get_acs(geography = "place", variables = "B19013_001",
  state = "CA", year = 2023) %>%
  filter(grepl("Stockton city", NAME)) %>% pull(estimate)
cat("Stockton city-wide median household income: $",
    formatC(st_city_median, format = "d", big.mark = ","), "\n\n")

# get tract-level income for San Joaquin county with geometry
st_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "San Joaquin", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild Stockton school buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE)
st_buf <- school_coords %>% filter(City == "Stockton") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# calculate weighted average income within each 1-mile buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  intersection <- st_intersection(tracts, buf); if (nrow(intersection) == 0) return(NA_real_)
  tract_areas   <- as.numeric(st_area(tracts))
  overlap_areas <- as.numeric(st_area(intersection))
  matched_areas <- tract_areas[match(intersection$GEOID, tracts$GEOID)]
  proportions   <- pmin(overlap_areas / matched_areas, 1)
  round(weighted.mean(intersection$median_income, proportions, na.rm = TRUE))
}

st_buf$median_income <- vapply(seq_len(nrow(st_buf)),
  function(i) income_in_buffer(st_buf[i, ], st_income_raw), numeric(1))

# pull tract-level race COUNTS with geometry (B03002 = Hispanic origin by race)
st_race_raw <- get_acs(geography = "tract",
  variables = c(total="B03002_001", white="B03002_003", black="B03002_004",
                amind="B03002_005", asian="B03002_006", pacisl="B03002_007",
                hispanic="B03002_012"),
  state = "CA", county = "San Joaquin",
  geometry = TRUE, year = 2023, output = "wide") %>%
  st_transform(crs = 3310)

# weighted racial % within each 1-mile buffer (areal interpolation)
race_in_buffer <- function(buf, race_sf) {
  hits <- st_intersects(race_sf, buf, sparse = FALSE)[, 1]
  tracts <- race_sf[hits, ]; if (nrow(tracts) == 0) return(rep(NA_real_, 7))
  inter <- st_intersection(tracts, buf); if (nrow(inter) == 0) return(rep(NA_real_, 7))
  ta <- as.numeric(st_area(tracts)); oa <- as.numeric(st_area(inter))
  prop <- pmin(oa / ta[match(inter$GEOID, tracts$GEOID)], 1)
  tot    <- sum(inter$totalE    * prop, na.rm = TRUE)
  black  <- sum(inter$blackE    * prop, na.rm = TRUE)
  hisp   <- sum(inter$hispanicE * prop, na.rm = TRUE)
  white  <- sum(inter$whiteE    * prop, na.rm = TRUE)
  asian  <- sum(inter$asianE    * prop, na.rm = TRUE)
  amind  <- sum(inter$amindE    * prop, na.rm = TRUE)
  pacisl <- sum(inter$pacislE   * prop, na.rm = TRUE)
  other  <- tot - black - hisp - white - asian - amind - pacisl
  if (tot == 0) return(rep(NA_real_, 7))
  round(c(black, hisp, white, asian, amind, pacisl, other) / tot * 100, 1)
}

race_mat <- t(vapply(seq_len(nrow(st_buf)),
  function(i) race_in_buffer(st_buf[i, ], st_race_raw), numeric(7)))
st_buf$`% Black`            <- paste0(race_mat[, 1], "%")
st_buf$`% Hispanic`         <- paste0(race_mat[, 2], "%")
st_buf$`% White`            <- paste0(race_mat[, 3], "%")
st_buf$`% Asian`            <- paste0(race_mat[, 4], "%")
st_buf$`% Native American`  <- paste0(race_mat[, 5], "%")
st_buf$`% Pacific Islander` <- paste0(race_mat[, 6], "%")
st_buf$`% Other`            <- paste0(race_mat[, 7], "%")

# join income + race onto existing st_data
st_full <- st_data %>%
  left_join(st_buf %>% st_drop_geometry() %>%
    select(School, median_income, `% Black`, `% Hispanic`, `% White`,
           `% Asian`, `% Native American`, `% Pacific Islander`, `% Other`),
    by = "School") %>%
  mutate(
    `Buffer Income`   = paste0("$", formatC(median_income, format = "d", big.mark = ",")),
    `vs. City Median` = ifelse(
      (median_income - st_city_median) >= 0,
      paste0("+$", formatC(median_income - st_city_median, format = "d", big.mark = ","), " above"),
      paste0("-$", formatC(abs(median_income - st_city_median), format = "d", big.mark = ","), " below"))
  )

# top 5 most at-risk
cat("--- Stockton: Top 5 Most At-Risk ---\n")
print(st_full %>% arrange(desc(risk_score)) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`,
         `% Black`, `% Hispanic`, `% White`, `% Asian`,
         `% Native American`, `% Pacific Islander`, `% Other`))

# top 5 least at-risk
cat("\n--- Stockton: Top 5 Least At-Risk ---\n")
print(st_full %>% arrange(risk_score) %>% head(5) %>%
  select(School, `Risk Score` = risk_score, `Buffer Income`, `vs. City Median`,
         `% Black`, `% Hispanic`, `% White`, `% Asian`,
         `% Native American`, `% Pacific Islander`, `% Other`))
```

FOR SAN JOSE: Scatterplot CAASPP vs. Risk
```{r SJ scatter CAASPP vs. Risk, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)

# one point per San Jose school: risk score (x) vs. CAASPP % unsatisfactory (y)
# drop schools missing a CAASPP score (e.g. suppressed data)
sj_scatter <- sj_data %>%
  filter(!is.na(caaspp_unsatisfactory))

ggplot(sj_scatter, aes(x = risk_score, y = caaspp_unsatisfactory * 100)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  # add a linear line of best fit
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    title = "Risk Score vs. CAASPP Unsatisfactory Rate — San Jose Public High Schools",
    x     = "Risk Score (0 = least risk, 1 = most risk)",
    y     = "CAASPP % Unsatisfactory"
  ) +
  theme_minimal()
```

FOR SAN JOSE: Scatterplot Risk & CAASPP vs. Income
```{r SJ scatter CAASPP & Risk vs. Income, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(tidyr)
library(scales)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get tract-level income for Santa Clara county
sj_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "Santa Clara", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild San Jose buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE) %>%
  filter(City == "San Jose")
sj_buf <- school_coords %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# weighted average income within each buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  inter <- st_intersection(tracts, buf); if (nrow(inter) == 0) return(NA_real_)
  ta <- as.numeric(st_area(tracts)); oa <- as.numeric(st_area(inter))
  prop <- pmin(oa / ta[match(inter$GEOID, tracts$GEOID)], 1)
  round(weighted.mean(inter$median_income, prop, na.rm = TRUE))
}
sj_buf$median_income <- vapply(seq_len(nrow(sj_buf)),
  function(i) income_in_buffer(sj_buf[i, ], sj_income_raw), numeric(1))

# join income to risk + CAASPP, drop schools missing CAASPP
sj_scatter2 <- sj_data %>%
  left_join(sj_buf %>% st_drop_geometry() %>% select(School, median_income), by = "School") %>%
  filter(!is.na(caaspp_unsatisfactory), !is.na(median_income)) %>%
  select(School, median_income,
         `Risk Score` = risk_score,
         `CAASPP Unsatisfactory` = caaspp_unsatisfactory) %>%
  pivot_longer(cols = c(`Risk Score`, `CAASPP Unsatisfactory`),
               names_to = "Outcome", values_to = "value")

# plot both outcomes vs. income, each with its own colored best-fit line
ggplot(sj_scatter2, aes(x = median_income, y = value, color = Outcome)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Risk Score" = "steelblue",
                                "CAASPP Unsatisfactory" = "darkorange")) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Risk Score & CAASPP Unsatisfactory vs. Median Household Income — San Jose",
    x     = "Median Household Income (lowest to highest)",
    y     = "Score (0.0 - 1.0)",
    color = "Outcome Variable"
  ) +
  theme_minimal()
```

FOR STOCKTON: Scatterplot CAASPP vs. Risk
```{r ST scatter CAASPP vs. Risk, echo = TRUE, eval = FALSE, message = FALSE} 
library(ggplot2)
library(dplyr)

# one point per Stockton school: risk score (x) vs. CAASPP % unsatisfactory (y)
# drop schools missing a CAASPP score (e.g. suppressed data)
st_scatter <- st_data %>%
  filter(!is.na(caaspp_unsatisfactory))

ggplot(st_scatter, aes(x = risk_score, y = caaspp_unsatisfactory * 100)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +
  # add a linear line of best fit
  geom_smooth(method = "lm", se = TRUE, color = "darkorange4") +
  labs(
    title = "Risk Score vs. CAASPP Unsatisfactory Rate — Stockton Public High Schools",
    x     = "Risk Score (0 = least risk, 1 = most risk)",
    y     = "CAASPP % Unsatisfactory"
  ) +
  theme_minimal()
```

FOR STOCKTON: Scatterplot Risk & CAASPP vs. Income
```{r ST scatter CAASPP & Risk vs. Income, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(tidyr)
library(scales)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# get tract-level income for San Joaquin county
st_income_raw <- get_acs(geography = "tract", variables = "B19013_001",
  state = "CA", county = "San Joaquin", geometry = TRUE, year = 2023) %>%
  st_transform(crs = 3310) %>% rename(median_income = estimate)

# rebuild Stockton buffers
school_coords <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE) %>%
  filter(City == "Stockton")
st_buf <- school_coords %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)

# weighted average income within each buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  inter <- st_intersection(tracts, buf); if (nrow(inter) == 0) return(NA_real_)
  ta <- as.numeric(st_area(tracts)); oa <- as.numeric(st_area(inter))
  prop <- pmin(oa / ta[match(inter$GEOID, tracts$GEOID)], 1)
  round(weighted.mean(inter$median_income, prop, na.rm = TRUE))
}
st_buf$median_income <- vapply(seq_len(nrow(st_buf)),
  function(i) income_in_buffer(st_buf[i, ], st_income_raw), numeric(1))

# join income to risk + CAASPP, drop schools missing CAASPP, reshape to long format
st_scatter2 <- st_data %>%
  left_join(st_buf %>% st_drop_geometry() %>% select(School, median_income), by = "School") %>%
  filter(!is.na(caaspp_unsatisfactory), !is.na(median_income)) %>%
  select(School, median_income,
         `Risk Score` = risk_score,
         `CAASPP Unsatisfactory` = caaspp_unsatisfactory) %>%
  pivot_longer(cols = c(`Risk Score`, `CAASPP Unsatisfactory`),
               names_to = "Outcome", values_to = "value")

# plot both outcomes vs income, each with its own colored best-fit line
ggplot(st_scatter2, aes(x = median_income, y = value, color = Outcome)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Risk Score" = "steelblue",
                                "CAASPP Unsatisfactory" = "darkorange")) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Risk Score & CAASPP Unsatisfactory vs. Median Household Income — Stockton",
    x     = "Median Household Income (lowest to highest)",
    y     = "Score (0.0 - 1.0)",
    color = "Outcome Variable"
  ) +
  theme_minimal()
```

FOR BOTH CITIES: Scatterplot CAASPP vs. Risk
```{r both cities scatter CAASPP vs. Risk, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)

# combine both cities (all 46 schools), drop schools missing CAASPP
both_scatter <- all_data %>%
  filter(!is.na(caaspp_unsatisfactory))

ggplot(both_scatter, aes(x = risk_score, y = caaspp_unsatisfactory * 100)) +
  # color the dots by city, but keep ONE shared best-fit line
  geom_point(aes(color = City), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(values = c("San Jose" = "steelblue",
                                "Stockton" = "darkorange")) +
  labs(
    title = "Risk Score vs. CAASPP Unsatisfactory Rate — All 46 Public High Schools",
    x     = "Risk Score (0 = least risk, 1 = most risk)",
    y     = "CAASPP % Unsatisfactory",
    color = "City"
  ) +
  theme_minimal()
```

Both cities: Linear regression model - CAASPP ~ Risk Score
```{r CAASPP ~ Risk Score linear regression, echo = TRUE, eval = FALSE, message = FALSE}
library(tidymodels)

# one row per school; drop schools missing a CAASPP score
sj_reg  <- sj_data  %>% filter(!is.na(caaspp_unsatisfactory))
st_reg  <- st_data  %>% filter(!is.na(caaspp_unsatisfactory))
all_reg <- all_data %>% filter(!is.na(caaspp_unsatisfactory))

# specify a linear regression model using the lm engine
lm_model      <- linear_reg() %>% set_engine("lm")
# fit metrics: rsq (closer to 1 = better fit) and rmse 
model_metrics <- metric_set(rsq, rmse)

# San Jose 
cat("\n--- SAN JOSE: CAASPP ~ Risk Score (n =", nrow(sj_reg), ") ---\n")
sj_fit <- lm_model %>% fit(caaspp_unsatisfactory ~ risk_score, data = sj_reg)
sj_fit                                            # intercept + slope
sj_fit %>%
  predict(sj_reg) %>%
  bind_cols(sj_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)     # rsq + rmse

# Stockton
cat("\n--- STOCKTON: CAASPP ~ Risk Score (n =", nrow(st_reg), ") ---\n")
st_fit <- lm_model %>% fit(caaspp_unsatisfactory ~ risk_score, data = st_reg)
st_fit
st_fit %>%
  predict(st_reg) %>%
  bind_cols(st_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)

# both
cat("\n--- BOTH CITIES: CAASPP ~ Risk Score (n =", nrow(all_reg), ") ---\n")
all_fit <- lm_model %>% fit(caaspp_unsatisfactory ~ risk_score, data = all_reg)
all_fit
all_fit %>%
  predict(all_reg) %>%
  bind_cols(all_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)
```

Both cities: Linear regression model - CAASPP ~ MAI & TSI
```{r CAASPP ~ MAI & TSI linear regression, echo = TRUE, eval = FALSE, message = FALSE}
library(tidymodels)

# one row per school; drop schools missing a CAASPP score
sj_reg  <- sj_data  %>% filter(!is.na(caaspp_unsatisfactory))
st_reg  <- st_data  %>% filter(!is.na(caaspp_unsatisfactory))
all_reg <- all_data %>% filter(!is.na(caaspp_unsatisfactory))

# specify a linear regression model using the lm engine 
lm_model      <- linear_reg() %>% set_engine("lm")
# fit metrics: rsq (closer to 1 = better fit) and rmse 
model_metrics <- metric_set(rsq, rmse)

# San Jose 
cat("\n--- SAN JOSE: CAASPP ~ MAI + TSI (n =", nrow(sj_reg), ") ---\n")
sj_fit2 <- lm_model %>%
  fit(caaspp_unsatisfactory ~ mobility_access_index + transportation_safety_index,
      data = sj_reg)
sj_fit2                                           # intercept + two slopes
sj_fit2 %>%
  predict(sj_reg) %>%
  bind_cols(sj_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)     # rsq + rmse

# Stockton
cat("\n--- STOCKTON: CAASPP ~ MAI + TSI (n =", nrow(st_reg), ") ---\n")
st_fit2 <- lm_model %>%
  fit(caaspp_unsatisfactory ~ mobility_access_index + transportation_safety_index,
      data = st_reg)
st_fit2
st_fit2 %>%
  predict(st_reg) %>%
  bind_cols(st_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)

# both cities
cat("\n--- BOTH CITIES: CAASPP ~ MAI + TSI (n =", nrow(all_reg), ") ---\n")
all_fit2 <- lm_model %>%
  fit(caaspp_unsatisfactory ~ mobility_access_index + transportation_safety_index,
      data = all_reg)
all_fit2
all_fit2 %>%
  predict(all_reg) %>%
  bind_cols(all_reg %>% select(caaspp_unsatisfactory)) %>%
  model_metrics(caaspp_unsatisfactory, .pred)
```

Both cities: Linear regression model - Risk ~ Income
```{r Risk ~ Income linear regression, echo = TRUE, eval = FALSE, message = FALSE}
library(tidycensus)
library(sf)
library(dplyr)
library(readr)
library(tidymodels)

DATA <- "~/Documents/SOC10_FINALPROJECT/Data Sets"
census_api_key("1aa7f6027937ad760c000edaceb8770891050792")

# area-weighted median income within one 1-mile buffer
income_in_buffer <- function(buf, income_sf) {
  hits <- st_intersects(income_sf, buf, sparse = FALSE)[, 1]
  tracts <- income_sf[hits, ]; if (nrow(tracts) == 0) return(NA_real_)
  inter <- st_intersection(tracts, buf); if (nrow(inter) == 0) return(NA_real_)
  ta <- as.numeric(st_area(tracts)); oa <- as.numeric(st_area(inter))
  prop <- pmin(oa / ta[match(inter$GEOID, tracts$GEOID)], 1)
  round(weighted.mean(inter$median_income, prop, na.rm = TRUE))
}

# get buffer income for one city
buffer_income <- function(city, county) {
  inc <- get_acs(geography = "tract", variables = "B19013_001",
                 state = "CA", county = county, geometry = TRUE, year = 2023) %>%
    st_transform(crs = 3310) %>% rename(median_income = estimate)
  buf <- read_csv(file.path(DATA, "school_coords.csv"), show_col_types = FALSE) %>%
    filter(City == city) %>%
    st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
    st_transform(crs = 3310) %>% st_buffer(dist = 1609.34)
  buf$median_income <- vapply(seq_len(nrow(buf)),
    function(i) income_in_buffer(buf[i, ], inc), numeric(1))
  buf %>% st_drop_geometry() %>% select(School, median_income)
}

sj_inc <- buffer_income("San Jose", "Santa Clara")
st_inc <- buffer_income("Stockton", "San Joaquin")

# join income onto risk scores; income in dollars
sj_risk_inc <- sj_data %>% left_join(sj_inc, by = "School") %>% filter(!is.na(median_income))
st_risk_inc <- st_data %>% left_join(st_inc, by = "School") %>% filter(!is.na(median_income))

# linear regression model using the lm engine 
lm_model      <- linear_reg() %>% set_engine("lm")
# fit metrics: rsq (closer to 1 = better fit) and rmse (Lecture 7)
model_metrics <- metric_set(rsq, rmse)

# SJ
cat("\n--- SAN JOSE: Risk Score ~ Median Income (n =", nrow(sj_risk_inc), ") ---\n")
sj_fit4 <- lm_model %>% fit(risk_score ~ median_income, data = sj_risk_inc)
sj_fit4                                           # intercept + slope (per $1)
sj_fit4 %>%
  predict(sj_risk_inc) %>%
  bind_cols(sj_risk_inc %>% select(risk_score)) %>%
  model_metrics(risk_score, .pred)                # rsq + rmse

# Stockton 
cat("\n--- STOCKTON: Risk Score ~ Median Income (n =", nrow(st_risk_inc), ") ---\n")
st_fit4 <- lm_model %>% fit(risk_score ~ median_income, data = st_risk_inc)
st_fit4
st_fit4 %>%
  predict(st_risk_inc) %>%
  bind_cols(st_risk_inc %>% select(risk_score)) %>%
  model_metrics(risk_score, .pred)
```

Scatterplot ~ Risk vs. Income
```{r scatterplot risk ~ income, echo = TRUE, eval = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(scales)

# NOTE: run the "Risk ~ Income linear regression" chunk above first

# combine both cities into one table for plotting
risk_income <- bind_rows(sj_risk_inc, st_risk_inc)

# one point per school: buffer median household income (x) vs. Risk Score (y),
# with a linear best-fit line and shaded confidence band (se = TRUE), one per city
ggplot(risk_income, aes(x = median_income, y = risk_score, color = City)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("San Jose" = "steelblue",
                                "Stockton" = "darkorange")) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title    = "Risk Score vs. Median Household Income — Public High Schools",
    subtitle = "Each point = one school's 1-mile buffer; line = linear regression fit",
    x        = "Buffer Median Household Income",
    y        = "Risk Score (0 = least risk, 1 = most risk)",
    color    = "City",
    caption  = "San Jose R-squared = 0.35  |  Stockton R-squared = 0.17  |  Source: 2019-2023 ACS"
  ) +
  theme_minimal()
```
